---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-", # Crucial: Saves plots to the figures folder
  out.width = "100%"
)
```

# pima R package

A package devoted to multivariate resampling-based tests. By resampling jointly on all univariate tests (e.i. sign-flip score tests by Hemerik, Goeman and Finos (2020)) it allows for multivariate and selective inference -- e.g. weak and strong control of the Familywise Error Rate or confidence bounds for True Discovery Proportion.


<!-- #[library flipscores on CRAN](http://cran.r-project.org/web/packages/flipscores/index.html) -->


* * *

## Set up

To **install** this github version type (in R):

    ##if devtools is not installed yet: 
    ## install.packages("devtools") 
    library(devtools)
    install_github("livioivil/pima")




## A (minimal) example

```{r,  message = FALSE}
devtools::load_all()
library(patchwork)
library(ggplot2)

n=20
set.seed(123)
D=data.frame(X=rnorm(n),Z1=rnorm(n),Z2=rnorm(n))
D$Y=D$Z1+D$X+rnorm(n)
D$Y2 = D$Z1+D$X+rnorm(n)
mod1=glm(Y~X+Z1+Z2,data=D)
mod2=glm(Y~X+poly(Z1,2)+Z2,data=D)
mod3=glm(Y~X+poly(Z1,2)+poly(Z2,2),data=D)
mod4=glm(Y~X+Z1+poly(Z2,2),data=D)
mod5= glm(Y2~X,data=D)
mods=list(mod1=mod1,mod2=mod2,mod3=mod3,mod4=mod4, mod5 = mod5)

res = pima(mods)

# by default the function create the coefficient name as
# response ~ coefficient
plot(res)

# you can also avoid this and divide the response variable
plot(res, facet = response ~ .)

# to create different specification curves 
plot(res, facet = response ~ ., which.response = "Y") /
  plot(res, facet = response ~ ., which.response = "Y2")

# this (mostly) works also for the specification curve

# by default create the combination of response/coefficient
spec_curve(res)

# subset a specific response
spec_curve(res, which.response = "Y")

# faceting the response
spec_curve(res, facet.y = TRUE)
```


## References

J Hemerik, JJ Goeman and L Finos (2019) Robust testing in generalized linear models by sign-flipping score contributions. Journal of the Royal Statistical Society Series B: Statistical Methodology, Volume 82, Issue 3, July 2020, Pages 841–864.  
<https://doi.org/10.1111/rssb.12369>

R De Santis, J Goeman, J Hemerik, L Finos (2022)
Inference in generalized linear models with robustness to misspecified variances arXiv: 2209.13918.  
<https://arxiv.org/abs/2209.13918>

P Girardi, A Vesely, D Lakens, G Altoè, M Pastore, A Calcagnì, L Finos (2022) Post-selection Inference in Multiverse Analysis (PIMA): an inferential framework based on the sign flipping score test. arxiv: 2210.02794.  
<https://arxiv.org/abs/2210.02794>


## Bug reports
If you encounter a bug, please file a
[reprex](https://github.com/tidyverse/reprex) (minimal reproducible example)
on [github](https://github.com/livioivil/pima/issues).
