---
title: "Hurricane dataset"
output: rmarkdown::html_vignette
author: "Vari ed eventuali"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Hurricane dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(pima)
devtools::load_all()
library(jointest)
library(flipscores)
# results reporting
library(gtsummary)
library(tidyverse)
library(GGally)
library(knitr)
library(gridExtra)
library(corrplot)
# models 
library(MASS)
# read STATA dataset
library(readstata13)
library(here)

opts_chunk$set(echo=FALSE,
               cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)

hurricane<-read.dta13(here("data", "hurricanes_2015_01_09.dta"))

#Original analysis from https://www.pnas.org/content/111/24/8782  
#year: year of the hurricane
#masfem: 1 masculine - 11 feminine
#min: minimum of pression
#alldeaths: total deaths
#ndam: amount of damages normalized 
#elapsedyrs: years since analysis
#masfem_mturk: femininity with mturk
#wind: highest wind speed
#ndam15: amount of damages normalized to the year 2015
```

# Hurricane  
## Dataset pre processing

```{r echo=TRUE}
hurricane$gender_mf<-as.factor(hurricane$gender_mf)
hurricane$category<-as.factor(hurricane$category)
levels(hurricane$gender_mf)<-c("Male","Female")
hurricane[,c(1,3,4,5,7,12,13)] %>% tbl_summary(by = gender_mf) %>% add_p() %>%as_gt() %>%   gt::as_latex()
#we consider only deaths

p1<-ggplot(hurricane, aes(x=gender_mf, y=alldeaths, color=gender_mf)) +
  geom_boxplot()
p2<-ggplot(hurricane, aes(x=gender_mf, y=ndam15, color=gender_mf)) +
  geom_boxplot()

grid.arrange(p1,p2,ncol=2)
# to export
#pdf("figure1.pdf",width = 10,height = 4)
#grid.arrange(p1,p2,ncol=2)
#dev.off()

# correlation plot
corrplot(cor(hurricane[,c("alldeaths","masfem","min","wind","ndam15","year")]))
hurricane$Log_deaths<-log(hurricane$alldeaths+1)

pm <- ggpairs(hurricane, columns = c(14,3,1,4,12,13),upper = list(continuous = wrap("cor", method = "spearman")))
pm
#to export
#pdf("corr.pdf",width=8,height =5)
#pm
#dev.off()
```

Correlation very high between min and wind (-0.85)

## GLM Model

Quasi-Poisson model since the number of deaths right-skew distributed  

```{r,echo=TRUE}
hist(hurricane$alldeaths,ylab="Frequency",xlab="Number of deaths")
```

and Var(Y) >> E[Y] (`r round(var(hurricane$alldeaths),1)` > `r round(mean(hurricane$alldeaths),1)`

Observational study design:
- Outcome: Number of Death
- Determinant: Femininity of the Hurricane Name
- Confounders: Year, Min of pressure, Wind, Amount of damages normalized at 2015. 
We consider the following glm:  

$$g(\mu)=\alpha+f_1(\mbox{mf})+f_2(\mbox{year})+f_3(\mbox{min})+f_4(\mbox{wind})+f_5(\mbox{ndam15})$$

where:  

$$\mbox{alldeaths} \sim quasiPoisson(\lambda=g^{-1}(X\beta),\phi)$$
where $\lambda$ is mean parameter and $\phi$ is estimated to deal with overdispersion (scale). Use set the $\log()$ as link function $g()$

Some considerations:  
- limited sample size: n=94  
- model with 5 predictors and 1 intercept  

We consider a simple additive regression model like the follwing one:

$$g(\mu)=\alpha+\beta_1 mf+ \beta_2 year+\beta_3 min+\beta_4 wind+\beta_5 ndam15$$
The coefficient $\beta_1$ expresses the increase of the log-death due to a 1-point increase in the $mf$ variable. The results of a Poisson regression models are commonly expressed in terms of Incidence Rate Ratio (IRR) just by exponentiation of the coefficients. (IRR=$\exp(\beta_1)$).

### Theoretical model  
We know in advance that the number of deaths depends on the strength of the hurricane (combination of min and wind) and of high population density and relative damages (ndam15).
In addition the year is an important confounder for two reasons:  
- 1) the alternating of the male/female name to an hurricane began in 1978 (https://www.nhc.noaa.gov/aboutnames_history.shtml)   
- 2) The year is a key factor for assessing the spread of information and its use in the general population over the year.   

```{r,echo=TRUE}
ggplot(hurricane,aes(x=year,y=masfem,col=gender_mf))+ylab("Degree of femininity")+geom_point(size=3)+xlab("Year")+labs(col="Gender")
```


## MULTIVERSE ANALYSIS 

All the four confounders are continuous. In addition to the identity, we decide to explore for each confounders the following transformations:   
- Square Root ($\sqrt(x)$);  
- Log ($\sqrt(x)$);  
- Second order polynomial ($x+x^2$).  

```{r,echo=TRUE}
########################################### possible specifications:
combinations_matrix=expand.grid(list(
  c("min","sqrt(min)","log(min)","poly(min,2)"),
  c("wind","sqrt(wind)","log(wind)","poly(wind,2)"),
  c("year","sqrt(year)","log(year)","poly(year,2)"),
  c("ndam15","sqrt(ndam15)","log(ndam15)","poly(ndam15,2)")))

combinations=apply(combinations_matrix,1,paste,collapse="+")
```

Then we combine all the transformations in $4^4=256$ specifications  

```{r,echo=TRUE}
##################### one list of formulas for each comparison of interest:
#######
formulas1=sapply(combinations,function(cmb) paste0("alldeaths ~ masfem +",cmb))
mod1=glm(as.formula(formulas1[1]),data=hurricane,family=quasipoisson())
mods1=lapply(formulas1, function(frml) update(mod1,formula. = as.formula(frml)))
```

Using the \texttt{pima} function() we apply the PIMA procedure to the list of models inside the mods1 list.

```{r,echo=TRUE}
res1=pima(mods1,tested_coeffs =  c("masfem"),n_flips = 5000,
                          score_type = "standardized",seed = 167)
names(res1)
plot.pima(res1,p.values="raw")
plot.pima(res1,p.values="adjusted")
```

The output is composed by:  
- Tspace: output table of sign_flip score testing  
- summary_table: overall results  
- mods: 256 models estimated (one for each specification)  
- info: specifications info (auxiliary object)  


### Specification curve 
```{r,echo=TRUE}
#######
#ggplot2 e cowplot
library(ggplot2)
library(cowplot)
spec_curve1<-spec_curve(res1,alpha = 0.05,p.values = "raw")
spec_curve1
spec_curve2<-spec_curve(res1,alpha = 0.05,p.values = "adjusted")
spec_curve2

t_stat_h0<-apply(res1$Tspace,1,max)
observed_t<-max(summary(res1)$Score)
2*mean(abs(t_stat_h0)>abs(observed_t))
ho_t<-ggplot(data = data.frame(t_stat_h0), aes(x = t_stat_h0)) +
  geom_histogram(aes(y = ..density..), fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = observed_t), color = "red", linetype = "dashed", size = 1.2) +
  geom_text(aes(x = observed_t, y = 0.00035, label = paste("Observed S =", observed_t)),
            color = "red", angle = 90, vjust = -0.5, hjust = -0.5, size = 4) +
  labs(title = "Histogram of Sign-flip test statistic Under H0",
       x = "S-statistic",
       y = "Density") +
  theme_minimal()

#pdf("null_dist.pdf",height = 5,width = 8)
ho_t
#dev.off()


#adjusted pvalues
res1$summary_table$p.adj<-jointest:::maxT.light(res1$Tspace,exp(seq(-8,0,0.5)))

p3<-plot(res1)
#pdf("graph_raw.pdf",height = 5,width = 8)
p3
#dev.off()

p4 <- plot(res1,p.values = "adjusted")
#pdf("graph_adj.pdf",height = 5,width = 8)
p4
#dev.off()

sum(res1$summary_table$p.adj<.05)

library(rpart)
library(rpart.plot)
pima_tree(res=res1) #raw
pima_tree(res=res1,p.values = "adjusted") #FWER adjusted pvalues
pima_tree(res=res1,p.values = "adjusted",cp=0.001) #higher complexity
```

### Pick up a model


Since the overall p-value is lower than <0.05, we can select among the estimated models a significant model 

```{r,echo=TRUE}
sel_model<-which.min(res1$summary_table$p.adj)
summary(mods1[[sel_model]])
exp(summary(mods1[[sel_model]])$coef[2,1])
```

In the selected model the adjusted effect of the masfem variable is significant. 1-point increase in masfem variables implies an increase of the risk of death of `r  round((exp(summary(mods1[[sel_model]])$coef[2,1])-1)*100,2)` \%