---
title: "PIMA workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PIMA workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

```{css, echo = FALSE}
.main-container {
  max-width: 1180px;
}
```

```{r setup}
library(pima)
library(ggplot2)

n_flips_vignette <- 1000

spec_top_theme <- function() {
  ggplot2::theme_bw(base_size = 10) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(size = 12)
    )
}

spec_bottom_theme <- function() {
  ggplot2::theme_bw(base_size = 8) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 7),
      strip.text.y = ggplot2::element_text(size = 7, angle = 0),
      panel.spacing.y = grid::unit(0.08, "lines")
    )
}
```

`pima` is designed for post-selection inference after a multiverse analysis:
fit several plausible models, test one or more focal coefficients jointly, and
then inspect how estimates and p-values behave across specifications.

The examples use `n_flips = 1000` and knitr caching, which is enough for a
stable workflow preview without making the vignette too slow. Use more flips
when the numerical result matters.

# A minimal multiverse as a list of fitted models

The most direct input to `pima()` is a named list of fitted models. Here we use
the package's Pima Indians diabetes data and test the coefficient for number of
pregnancies across a few logistic regressions.

```{r list-models}
data("pimads")

pima_data <- na.omit(pimads)
pima_data$diabetes01 <- as.integer(pima_data$diabetes == "pos")

mods <- list(
  full = glm(
    diabetes01 ~ npreg + glucose + pressure + bmi + age,
    data = pima_data,
    family = binomial()
  ),
  compact = glm(
    diabetes01 ~ npreg + glucose + bmi,
    data = pima_data,
    family = binomial()
  ),
  sqrt_bmi = glm(
    diabetes01 ~ npreg + glucose + sqrt(bmi) + age,
    data = pima_data,
    family = binomial()
  ),
  glucose_only = glm(
    diabetes01 ~ npreg + glucose,
    data = pima_data,
    family = binomial()
  )
)

manual_res <- pima(
  mods,
  tested_coeffs = "npreg",
  n_flips = n_flips_vignette,
  method = "maxT",
  score_type = "standardized",
  seed = 1
)

summary(manual_res, digits = 3)
```

The main arguments are:

- `mods`: a list of model objects, usually `glm` objects.
- `tested_coeffs`: coefficients to test. It can be a character vector common to
  all models, or a list with one character vector per model.
- `n_flips`: number of sign flips.
- `method`: multiplicity adjustment, `"maxT"`, `"minP"`, or `"none"`.
- `score_type`, `seed`, `output_models`, and other arguments are passed to
  `flipscores::flipscores()`.

If the same conceptual effect has different coefficient names across models,
use a list for `tested_coeffs`.

```{r coefficient-list}
mods_mixed <- list(
  linear = glm(
    diabetes01 ~ npreg + glucose + bmi,
    data = pima_data,
    family = binomial()
  ),
  quadratic = glm(
    diabetes01 ~ I(npreg^2) + glucose + bmi,
    data = pima_data,
    family = binomial()
  )
)

mixed_res <- pima(
  mods_mixed,
  tested_coeffs = list(
    linear = "npreg",
    quadratic = "I(npreg^2)"
  ),
  n_flips = n_flips_vignette,
  seed = 1
)

summary(mixed_res, digits = 3)
```

# Adding specification metadata

When models are supplied manually, `pima()` tries to reconstruct the formula
decisions from the fitted models. You can add your own scenario-level metadata
with `extra`; this is useful for plotting or grouping by analytical decisions.

```{r extra}
extra <- data.frame(
  model = names(mods),
  adjustment = c("full", "compact", "sqrt_bmi", "minimal"),
  includes_age = c(TRUE, FALSE, TRUE, FALSE)
)

manual_res <- pima(
  mods,
  tested_coeffs = "npreg",
  n_flips = n_flips_vignette,
  seed = 1,
  extra = extra
)

manual_res
```

# Plotting a PIMA result

`plot()` creates a volcano-style view: an estimate-like quantity on the x-axis
and raw or adjusted p-values on the y-axis. The p-value transformation can be
`"z"`, `"-log10"`, `"raw"`, or a custom function.

```{r plot-basic, fig.width = 8, fig.height = 5}
plot(
  manual_res,
  focal = "npreg",
  xvar = "estimate",
  p.adjusted = TRUE,
  p.transf = "-log10"
)
```

The same plot can be tuned with the arguments users usually need in practice:

- `xvar`: any numeric column in `summary(res)`, for example `"estimate"`,
  `"pcor"`, `"score"` or `"z"`;
- `p.adjusted`: `TRUE` for adjusted p-values, `FALSE` for raw p-values;
- `p.transf`: `"z"`, `"-log10"`, `"raw"` or a custom function;
- `by`: grouping variable, including metadata supplied through `extra`;
- `facet`, `facet.scales`, `shapes`, `alpha`, `xlab`, and `ylab`.

```{r plot-by-extra, fig.width = 8, fig.height = 5}
plot(
  manual_res,
  by = "adjustment",
  xvar = "pcor",
  p.adjusted = FALSE,
  p.transf = "z",
  alpha = 0.05,
  shapes = c(1, 16),
  xlab = "Partial correlation",
  ylab = "Raw p-value, z scale"
)
```

Regular expressions are handy when coefficient names are generated by
transformations or factor contrasts.

```{r plot-regex, fig.width = 8, fig.height = 5}
plot(
  mixed_res,
  focal = "npreg",
  regex = TRUE,
  xvar = "estimate",
  p.adjusted = FALSE,
  p.transf = "raw",
  alpha = 0.10,
  facet.scales = "free"
)
```

`volcano()` adds quantile reference lines and p-value cutoffs on top of the same
plotting idea.

```{r volcano, fig.width = 8, fig.height = 5}
volcano(
  manual_res,
  focal = "npreg",
  p.adjusted = FALSE,
  p.cut = c(0.10, 0.05, 0.01),
  q = c(0.10, 0.50, 0.90)
)
```

# Specification curves

`spec_curve()` combines two panels: the top panel shows the result for each
specification, and the bottom panel shows the analytical decisions that generated
that specification.

```{r spec-curve-manual, fig.width = 10, fig.height = 7}
spec_curve(
  manual_res,
  focal = "npreg",
  yvar = "estimate",
  p.adjusted = TRUE,
  alpha = 0.05,
  tbr = c(0.55, 0.45),
  colors = c("grey45", "#0072B2"),
  shapes = c(1, 16),
  title = "Manual multiverse: coefficient estimates",
  xlab = "Model specification",
  ylab = "Log-odds estimate",
  top.theme = spec_top_theme,
  bottom.theme = spec_bottom_theme,
  conf.int = TRUE,
  redundant = FALSE
)
```

If you test several coefficients or several responses, `spec_curve()` can facet
the result.

```{r multi-response, fig.width = 10, fig.height = 8}
set.seed(123)
n <- 30
sim_data <- data.frame(
  x = rnorm(n),
  z1 = rnorm(n),
  z2 = rnorm(n)
)
sim_data$y1 <- sim_data$x + sim_data$z1 + rnorm(n)
sim_data$y2 <- sim_data$x + sim_data$z2 + rnorm(n)

sim_mods <- list(
  y1_linear = glm(y1 ~ x + z1 + z2, data = sim_data),
  y1_poly = glm(y1 ~ x + poly(z1, 2) + z2, data = sim_data),
  y2_linear = glm(y2 ~ x + z1 + z2, data = sim_data),
  y2_poly = glm(y2 ~ x + z1 + poly(z2, 2), data = sim_data)
)

sim_res <- pima(
  sim_mods,
  tested_coeffs = "x",
  n_flips = n_flips_vignette,
  seed = 1
)

plot(sim_res, facet = response ~ ., p.adjusted = FALSE)

spec_curve(
  sim_res,
  focal = "x",
  p.adjusted = FALSE,
  facet.y = TRUE,
  tbr = c(0.60, 0.40),
  which.response = c("y1", "y2"),
  title = "Two responses, one focal coefficient",
  top.theme = spec_top_theme,
  bottom.theme = spec_bottom_theme,
  redundant = FALSE
)
```

# Creating a multiverse with `create_multi()`

`create_multi()` builds the scenario grid for you. It crosses:

- predictor inclusion and transformations;
- optional row subsets;
- optional fitting specifications.

The right-hand side of `formula` must contain bare additive predictors. Put
transformations in `nfuns` or `cfuns`, not directly inside the formula.

```{r create-multi}
multi <- create_multi(
  diabetes01 ~ npreg + glucose + bmi + age,
  data = pima_data,
  focal = "npreg",
  nfuns = list(
    glucose = "sqrt",
    bmi = "sqrt",
    age = "sqrt"
  ),
  subset = list(
    high_pressure = pima_data$pressure > median(pima_data$pressure)
  ),
  fit.specs = list(
    logit = create_fit(
      stats::glm,
      family = stats::binomial()
    ),
    probit = create_fit(
      stats::glm,
      family = stats::binomial(link = "probit")
    )
  )
)

multi

head(multi$scenarios)
```

The `spec_<variable>` columns record the analytical choice for each predictor:
`"excluded"`, `"identity"`, or the transformation name.

Models are fitted when `fit.specs` is supplied. Failed fits are kept as error
objects and marked in `scenarios$fit_ok`, so it is usually convenient to subset
before calling `pima()`.

```{r create-multi-pima}
multi_ok <- subset(multi, fit_ok)

multi_res <- pima(
  multi_ok,
  tested_coeffs = "npreg",
  n_flips = n_flips_vignette,
  seed = 1
)

head(summary(multi_res, digits = 3))
```

The metadata from `create_multi()` is carried into the `pima` result. For a
large grid, the scatter plot is useful as an overview.

```{r create-multi-plots, fig.width = 9, fig.height = 5}
plot(
  multi_res,
  by = c("subset", "fit_spec"),
  xvar = "pcor",
  p.adjusted = FALSE
)
```

A full specification curve with more than one hundred scenarios is usually too
dense for a vignette page. A good reporting pattern is to show the complete
overview above and then draw focused curves for meaningful slices of the
multiverse.

```{r create-multi-focused, fig.width = 12, fig.height = 8}
multi_logit_all <- subset(
  multi_ok,
  subset == "all" & fit_spec == "logit"
)

multi_logit_res <- pima(
  multi_logit_all,
  tested_coeffs = "npreg",
  n_flips = n_flips_vignette,
  seed = 1
)

spec_curve(
  multi_logit_res,
  focal = "npreg",
  yvar = "pcor",
  p.adjusted = FALSE,
  alpha = 0.05,
  tbr = c(0.62, 0.38),
  colors = c("grey55", "#D55E00"),
  shapes = c(1, 16),
  title = "Focused specification curve: all rows, logit models",
  ylab = "Partial correlation",
  top.theme = spec_top_theme,
  bottom.theme = spec_bottom_theme,
  redundant = FALSE
)
```

When you really want to inspect every scenario, export the complete curve as a
standalone SVG/PDF instead of forcing it into the vignette body. The code below
is not evaluated here because a 108-scenario curve is too dense for inline
reading.

```{r create-multi-full-spec-curve, eval = FALSE}
full_curve <- spec_curve(
  multi_res,
  focal = "npreg",
  yvar = "pcor",
  p.adjusted = FALSE,
  tbr = c(0.55, 0.45),
  title = "Full specification curve",
  top.theme = spec_top_theme,
  bottom.theme = spec_bottom_theme,
  redundant = FALSE
)

ggplot2::ggsave(
  "multi-full-spec-curve.svg",
  full_curve,
  width = 16,
  height = 10
)
```

`transf.focal = TRUE` allows transformations of focal predictors too. This is
useful when the focal effect itself can be parameterized in multiple ways.

```{r transformed-focal}
multi_focal <- create_multi(
  diabetes01 ~ npreg + glucose + bmi,
  data = pima_data,
  focal = "npreg",
  nfuns = list(
    npreg = "sqrt",
    glucose = "sqrt",
    bmi = "sqrt"
  ),
  transf.focal = TRUE,
  fit.specs = list(
    logit = create_fit(
      stats::glm,
      family = stats::binomial()
    )
  )
)

head(multi_focal$formula_specs)
```

# Using the `multiverse` package

`pima()` also has experimental support for objects from the `multiverse`
package. The key convention is that each universe must create a fitted model
called `fit`.

This example uses the `hurricane` data shipped by `multiverse`, not the
`hurricane` data shipped by `pima`.

```{r multiverse-package, eval = requireNamespace("multiverse", quietly = TRUE)}
library(multiverse)

data("hurricane", package = "multiverse")

hurricane_mv <- transform(
  hurricane,
  log_NDAM = log(NDAM + 1)
)

M <- multiverse()

inside(M, {
  fit <- glm(
    alldeaths ~
      MasFem +
      branch(
        pressure,
        "original" ~ MinPressure_before,
        "updated" ~ Minpressure_Updated_2014
      ) +
      branch(
        damage,
        "raw" ~ NDAM,
        "logged" ~ log_NDAM
      ),
    data = hurricane_mv,
    family = quasipoisson()
  )
})

execute_multiverse(M)

parameters(M)

hurricane_res <- pima(
  M,
  tested_coeffs = "MasFem",
  n_flips = n_flips_vignette,
  seed = 1
)

summary(hurricane_res, digits = 3)
```

The branches declared in `multiverse::branch()` are stored as metadata and can
be used for grouping and specification curves.

```{r multiverse-package-plots, eval = requireNamespace("multiverse", quietly = TRUE), fig.width = 10, fig.height = 7}
plot(
  hurricane_res,
  by = c("pressure", "damage"),
  xvar = "estimate",
  p.transf = "-log10",
  p.adjusted = FALSE
)

spec_curve(
  hurricane_res,
  focal = "MasFem",
  p.adjusted = FALSE,
  tbr = c(0.60, 0.40),
  colors = c("grey55", "#009E73"),
  title = "Hurricane multiverse from the multiverse package",
  top.theme = spec_top_theme,
  bottom.theme = spec_bottom_theme,
  redundant = FALSE
)
```

# Adjusting and combining tests

`pima()` applies multiplicity adjustment during the main call when
`method = "maxT"` or `method = "minP"`. You can also compute raw p-values first
with `method = "none"` and then call `p.adjust()` later.

The next example uses a factor predictor, because it also lets us show
`combine_contrasts()`.

```{r adjustment-combinations-data}
set.seed(123)
n <- 40
factor_data <- data.frame(
  x = rnorm(n),
  z1 = rnorm(n),
  z2 = rnorm(n),
  group = factor(sample(c("A", "B", "C"), n, replace = TRUE))
)
factor_data$y <- 0.5 * factor_data$x +
  ifelse(factor_data$group == "C", 1.2, 0) +
  factor_data$z1 +
  rnorm(n)

factor_mods <- list(
  linear = glm(y ~ x + group + z1 + z2, data = factor_data),
  poly_z1 = glm(y ~ x + group + poly(z1, 2) + z2, data = factor_data),
  compact = glm(y ~ x + group + z1, data = factor_data)
)

raw_factor_res <- pima(
  factor_mods,
  tested_coeffs = c("x", "groupB", "groupC"),
  n_flips = n_flips_vignette,
  method = "none",
  seed = 1
)

head(summary(raw_factor_res, digits = 3))
```

Because `raw_factor_res` was computed with `method = "none"`, it contains raw
p-values only. The re-exported `p.adjust()` function adds adjusted p-values from
the same joint resampling distribution.

```{r p-adjust}
maxT_factor_res <- p.adjust(
  raw_factor_res,
  method = "maxT"
)

minP_factor_res <- p.adjust(
  raw_factor_res,
  method = "minP"
)

head(summary(maxT_factor_res, digits = 3))
head(summary(minP_factor_res, digits = 3))
```

`combine_tests()` answers global questions by combining test statistics. With no
`by` argument, it asks whether at least one tested effect is present anywhere in
the multiverse. With `by = "model"` or `by = "coefficient"`, it asks the same
question within each model or coefficient.

```{r combine-tests}
summary(combine_tests(maxT_factor_res))

summary(combine_tests(
  maxT_factor_res,
  by = "model"
))

summary(combine_tests(
  maxT_factor_res,
  by = "coefficient"
))
```

You can also define custom groups with `by_list`. The list elements can contain
test positions, or column names from `Tspace`, when you want groups that do not
correspond to a single metadata column.

```{r combine-tests-custom}
custom_groups <- list(
  focal_x = which(maxT_factor_res$summary_table$coefficient == "x"),
  factor_group = which(maxT_factor_res$summary_table$coefficient != "x")
)

summary(combine_tests(
  maxT_factor_res,
  by_list = custom_groups
))
```

`combine_contrasts()` combines the contrasts of factor variables into one global
test for the factor. In this example `groupB` and `groupC` become a single
`group` test within each model.

```{r combine-contrasts}
factor_global_res <- combine_contrasts(maxT_factor_res)

summary(factor_global_res)

summary(combine_tests(
  factor_global_res,
  by = "coefficient"
))
```

# Variation of estimates

`voe()` summarizes the variation of estimates and p-values across the
multiverse. `voe_variance()` decomposes variation in a selected estimate-like
column over the specification decisions recorded in the `pima` object.

```{r voe}
voe(
  multi_res,
  xvar = "pcor",
  p.adjusted = FALSE
)

voe_variance(
  multi_res,
  estimate = "pcor"
)$dec
```

# A tree over specification decisions

`pima_tree()` fits an `rpart` tree using the scenario metadata as predictors.
With `method = "class"`, the response is whether a p-value is below `alpha`.
With `method = "anova"`, the response is the p-value itself.

```{r pima-tree}
pima_tree(
  multi_res,
  p.values = "raw",
  method = "anova"
)
```

# Practical checklist

For a typical workflow:

1. Define the universe of reasonable analytical choices.
2. Fit it manually, with `create_multi()`, or with `multiverse`.
3. Call `pima()` with the focal coefficients and a real number of flips.
4. Inspect `summary()`, `plot()`, `spec_curve()`, and, when useful, `voe()` or
   `pima_tree()`.
5. Report the specification decisions and whether the conclusions depend on
   those decisions.
