---
title: "Tutorial: Hurricane dataset analysis"
author: "Vari ed eventuali"
date: "`r Sys.Date()`"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{rotating}
- \usepackage{longtable}
output:
  html_document:
  highlight: haddock
latex_engine: xelatex
toc: yes
---
```{r knitr_init, echo=TRUE, cache=FALSE}
#carico library
rm(list=ls())
library(gtsummary)
library(tidyverse)
library(GGally)
library(MASS)
library(knitr)
library(readstata13)
library(gridExtra)
opts_chunk$set(echo=FALSE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
hurricane<-read.dta13("~/Dropbox/Convegni/2024_ASA/hurricanes_2015_01_09.dta")
source("Rutils.R")
head(hurricane)

#Original analysis from  

# https://www.pnas.org/content/111/24/8782  
#year: anno
#masfem: 1 masculine, 11 feminile
#min: minimo di pressione
#categoria uragano
#alldeaths: decessi totali
#ndam: indice normalizzato di danni
#elapsedyrs:anni passati dal 2013
#masfem_mturk: femminilità media assegnata da amazon turk
#wind; vento 
#ndam15: indice normalizzato di danni al 2015
```
# Hurricane  
## Preparazione del dataset 

```{r echo=TRUE}
hurricane$gender_mf<-as.factor(hurricane$gender_mf)
hurricane$category<-as.factor(hurricane$category)
levels(hurricane$gender_mf)<-c("Male","Female")
hurricane[,c(1,3,4,5,7,12,13)] %>% tbl_summary(by = gender_mf) %>% add_p() %>%as_gt() %>%   gt::as_latex()
#we consider only deaths

p1<-ggplot(hurricane, aes(x=gender_mf, y=alldeaths, color=gender_mf)) +
  geom_boxplot()
p2<-ggplot(hurricane, aes(x=gender_mf, y=ndam15, color=gender_mf)) +
  geom_boxplot()

grid.arrange(p1,p2,ncol=2)
pdf("figure1.pdf",width = 10,height = 4)
grid.arrange(p1,p2,ncol=2)
dev.off()

corrplot::corrplot(cor(hurricane[,c("alldeaths","masfem","min","wind","ndam15","year")]))
hurricane$Log_deaths<-log(hurricane$alldeaths+1)
pm <- ggpairs(hurricane, columns = c(14,3,1,4,12,13),upper = list(continuous = wrap("cor", method = "spearman")))
pm
pdf("corr.pdf",width=8,height =5)
pm
dev.off()


```

Correlation very high between:  
- min and wind (-0.85)



## GLM Model

Quasi-Poisson model since the number of deaths right-skew distributed  

```{r,echo=TRUE}
hist(hurricane$alldeaths,ylab="Frequency",xlab="Number of deaths")
pdf("gender_year.pdf",width = 8,height = 4)
ggplot(hurricane,aes(x=year,y=masfem,col=gender_mf))+ylab("Degree of femininity")+geom_point(size=3)+xlab("Year")+labs(col="Gender")
dev.off()
```

and Var(Y) >> E[Y] (`r round(var(hurricane$alldeaths),1)` > `r round(mean(hurricane$alldeaths),1)`

We consider the following glm:  

$$g(\mu)=\alpha+f_1(\mbox{mf})+f_2(\mbox{year})+f_3(\mbox{min})+f_4(\mbox{elapsedyrs})+f_5(\mbox{wind})+f_6(\mbox{ndam15})$$

where:  

$$\mbox{alldeaths} \sim quasiPoisson(\lambda=g^{-1}(X\beta))$$
Some considerations:
- limited sample size: n=94
- model with 4 predictors and 1 intercept

### Theoretical model
We know in advance that the number of deaths depends on the strength of the hurricane (min,pressure,wind), if it touches a populated area (ndam15), and the historical period (year).


## MULTIVERSE ANALYSIS 

```{r,echo=TRUE}
########################################### possible specifications:
combinations_matrix=expand.grid(list(
  c("min","sqrt(min)","log(min)","poly(min,2)"),
  c("wind","sqrt(wind)","log(wind)","poly(wind,2)"),
  c("year","sqrt(year)","log(year)","poly(year,2)"),
  c("ndam15","sqrt(ndam15)","log(ndam15)","poly(ndam15,2)")))

combinations=apply(combinations_matrix,1,paste,collapse="+")
##################### one list of formulas for each comparison of interest:

#######
formulas1=sapply(combinations,function(cmb) paste0("alldeaths ~ masfem +",cmb))
mod1=glm(as.formula(formulas1[1]),data=hurricane,family=quasipoisson())
mods1=lapply(formulas1, function(frml) update(mod1,formula. = as.formula(frml)))

library(jointest)
library(flipscores)
res1=join_flipscores(mods1,n_flips = 5000,
                          tested_coeffs =  c("masfem"),
                          score_type = "standardized" ,seed = 123)
#dipendenze ggplot2 e cowplot
#3 argomenti 
# 1: res: results della funzione join_flipscores
# 2: alpha: valore per gli intervalli di confidenza 
p<-spec_curve(res1,alpha = 0.05)
#pdf("spec_curve.pdf",width = 8,height = 5)
p
#dev.off()

2*mean(apply(res1$Tspace,1,max)>max(summary(res1)$Score))
t_stat_h0<-apply(res1$Tspace,1,max)
observed_t<-max(summary(res1)$Score)
ho_t<-ggplot(data = data.frame(t_stat_h0), aes(x = t_stat_h0)) +
  geom_histogram(aes(y = ..density..), fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = observed_t), color = "red", linetype = "dashed", size = 1.2) +
  geom_text(aes(x = observed_t, y = 0.00035, label = paste("Observed S =", observed_t)),
            color = "red", angle = 90, vjust = -0.5, hjust = -0.5, size = 4) +
  labs(title = "Histogram of Sign-flip test statistic Under H0",
       x = "S-statistic",
       y = "Density") +
  theme_minimal()

#pdf("null_dist.pdf",height = 5,width = 8)
ho_t
#dev.off()


#adjusted pvalues
res1$summary_table$p.adj<-maxT.light(res1$Tspace,c(0.001,0.005,0.01,0.025,0.05,0.075,0.1,0.2,0.3,0.4,0.5))

p3<-plot(res1)
#pdf("graph_raw.pdf",height = 5,width = 8)
p3
#dev.off()

p4 <- plot(res1,p.values = "adjusted")
#pdf("graph_adj.pdf",height = 5,width = 8)
p4
#dev.off()

sum(res1$summary_table$p.adj<.05)

pvalue_tree(res=res1) # raw
pvalue_tree(res=res1,p.values = "adjusted") #FWER adjusted

#vibration of effects: statistiche descritive? 
#https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4555355/
#RHR 99-01% dei coefficienti, RP: 99-01% dei -log10(pvalue)
# Relative Incidence Rate Ratio -> Relative Coefficient Ratio
IRR<-exp(res1$summary_table$Estimate)
RIRR<-quantile(IRR,c(0.99))/quantile(IRR,c(0.01))
RIRR
#Relative pvalues
RP_raw<-diff(-log10(quantile(res1$summary_table$p,c(0.99,0.01))))
RP_raw
RP_adjusted<-diff(-log10(quantile(res1$summary_table$p.adj,c(0.99,0.01))))
RP_adjusted
```
